---
title: "Adding external data to ggseg plotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding external data to ggseg plotting}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE)
```

## Introduction
Once you have covered the main functionality in `ggseg` you will want to use it to plot the results of your data. 
In order to do this, your data must adhere to certain specifications, so that `ggseg` can manage to merge your data with the atlas you are using.
This means you need to be able to inspect and locate the way the regions you are working with are names in the internal atlas files.
This vignette should provide the tools you need to figure these features out, and to manipulate your data to fit these requirements.

## Inspecting the atlas labels
There are several ways you can inspect what the data in the atlas looks like.
While each atlas has some small differences, they all share six main columns:
**1. long** - x-axis  
**2. lat **- y-axis  
**3. area **- name of area/network  
**4. hemi** - hemisphere (left or right)  
**5. side** - side of view (medial, lateral, sagittal or axial)  

Most atlases also have a `label` column, which are raw names assigned from the program run to segment/extract data.

This information is stored in a list of data.frames called `atlas.info`, which is loaded when ggseg is loaded, just like the atlases and palettes.

```{r start}
library(ggseg)
library(magrittr)
library(dplyr)
atlas.info$yeo7
```

Here you can see information about the `yeo7` atlas, and the main attributes of this atlas. 
If you want to use external data with your `ggseg` plot, you will need to make sure that your data has at least one column corresponding in name and content with another in the atlas you are using. 

## Structuring data for merging
For instance, here we make some data for the "default" and "visual" networks in the `yeo7` atlas, and two p values for those two networks.

```{r}
someData = data.frame(area=c("default","visual"),
                      p=c(.03,.6), 
                      stringsAsFactors = F)
someData
```

Notice you we have spelled bothe the column name and the area names **exactly** as they appear in the data. 
This is necessary for the merging within the `ggseg` function to work properly.
This merge can be attempted before supplying the data to `ggseg` to see if there are any errors.

```{r}
yeo7 %>% 
  left_join(someData) %>% 
  head(10) #only added to truncate output
```

No errors!
Yes, the `p` column is seemingly full of `NA`s, but that is just because the top of the data is the somatomotor network, which we did not supply any p values for, so it has been populated with `NA`s.
We can sort the data differently, so we can see the `p`has been added correctly.

```{r}
yeo7 %>% 
  left_join(someData) %>% 
  arrange(p) %>% 
  head(10) #only added to truncate output
```

If you need your data to be matched on several columns, the approach is the same. 
Add the column you want to match on, with the **exact** same name, and make sure it's content matches the content of the same column in the data.

```{r}
someData$hemi = c("left","left")
someData

yeo7 %>% 
  left_join(someData) %>% 
  arrange(p) %>% 
  head(10)
```

Notice how the message now states that it is joining `by = c("area", "hemi")`.
The merge function has recognized that there are two equally named columns, and assumes (in this case correctly) that these are equivalent.  
**Notice** that everything is case-sensitive, so writing `Area` or `Left` will not result in matching.

## Providing data to `ggseg`
When you have managed to create data that merges nicely with the atlas, you can go ahead and supply it to the function.

```{r}
library(ggplot2)
ggseg(someData, atlas="yeo7", mapping=aes(fill=p))
```

You can actually also supply it directly as an atlas. 
For instance, if you had saved the merged data from the previous steps, you can supply this directly to the `atlas` option.

```{r}
newAtlas = yeo7 %>% 
  left_join(someData)

ggseg(atlas=newAtlas, mapping=aes(fill=p), position="stacked")
```

