---
title: "Using atlases with geom_sf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using atlases with geom_sf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6
)
```


While the new workflow in *ggseg* using the simple features type `geom_brain` works well, it struggles with integration with other very useful simple feature geoms, like labels. 
Since `geom_brain` can alter the position of slices through the `position` argument, and since the atlas is provided as a separate argument rather than integrated in the data itself, the geoms for sf-labels are not possible to implement. 
We also found it too tricky, with too many re-iterations of the same arguments, to implement a specialized brain-label geom. 

Because of these issues, we here provide an alternate workflow for those users wanting to implement other simple-feature geoms to the ggseg-plots. 
This workflow means not using the `geom_brain` function, but pre-joining and fixing the data _before_ providing it to `ggplot`. 
In many ways, this workflow mimics what happens behind the scenes in `ggseg()` and within the `geom_brain` functions.


# Explore the atlas

Firstly, you need to make sure that the atlas you are working on had simple features (sf) geometry column in it. 
The easiest way to do this is to simply call the atlas in the R terminal and look for information on `use` in the console printout.
The dk-atlas has `use: ggplot() + geom_brain()` letting us know that it is set up to work with simple features.

```{r setup}
library(ggseg)
library(ggplot2)

dk
```

You can have a look at the atlas and make sure it is what you expect

```{r}
plot(dk)
```

If you take an even closer look, you will see that the brain-atlas class is actually a list of four, where the `data` contains a simple features geometry column.
```{r}
dk$atlas
dk$type
dk$palette
dk$data
```


## If the atlas is not a simple-features version
We can force the dk brain-atlas into the original ggseg-atlas class and see how it looks different when the type is not set up with simple-features.

```{r}
dk2 <- as_ggseg_atlas(dk)
dk2
```

If the ggseg-atlas is set up correctly, we should be able to convert it into a simple-features atlas.

```{r}
dk2 <- as_brain_atlas(dk2)
dk2
```


# Joining the atlas with data

In most cases, users want to plot their own data projected onto the brain. 
In this workflow, you will need to get all the data, including having joined the atlas data, _before_ calling `ggplot`. 
This will then become the same workflow as any other simple-featured plotting, and many new possibilities of specialized sf-geoms opens to you. 

First, let's make up some data that we want to plot. Here a data.frame of 4 brain regions, with p-values connected to them.

```{r}
library(dplyr)

some_data <- tibble(
  region = c(
    "transverse temporal",
    "insula",
    "precentral",
    "superior parietal"
  ),
  p = sample(seq(0, .5, .001), 4)
)

some_data

```

To join a data.frame to an atlas object, you should use the specialized `brain_join` function. 
This makes sure that atlas information is preserved while joining. 

```{r}
some_data |>
  brain_join(dk)
```

The output is a standard sf-object, and can therefore also be plotted as one. 

```{r}
some_data |>
  brain_join(dk) |>
  plot()
```

To use ggplot for plotting the new sf-object, use standard ggplot-syntax with the `geom_sf` geom.

```{r}
some_data |>
  brain_join(dk) |>
  ggplot() +
  geom_sf(aes(fill = p))
```


If you want to reposition the slices, so that they are more to you liking, use the `reposition_brain` function (not to be confused with the ggproto-function `position_brain`!)
```{r}
some_data |>
  brain_join(dk) |>
  reposition_brain(hemi ~ side) |>
  ggplot() +
  geom_sf(aes(fill = p))
```


Once your data is plotted as a geom_sf, you can add geoms like `geom_sf_label` to add labels to the regions.

```{r}
some_data |>
  brain_join(dk) |>
  reposition_brain(hemi ~ side) |>
  ggplot(aes(fill = p)) +
  geom_sf(show.legend = FALSE) +
  geom_sf_label(
    aes(label = ifelse(!is.na(p), region, NA)),
    alpha = .8,
    show.legend = FALSE
  )
```

If you want a version where the labels do not overlap each other, you can try the `geom_sf_label_repel` function from the `ggrepel` package. 

## Faceting groups

If your data includes groups, and you want to facet the output, you need to group the data _before_ calling `brain_join`.

```{r}
some_data <- tibble(
  region = rep(
    c("transverse temporal", "insula", "precentral", "superior parietal"),
    2
  ),
  p = sample(seq(0, .5, .001), 8),
  groups = c(rep("g1", 4), rep("g2", 4))
)
some_data
```


```{r}
some_data |>
  group_by(groups) |>
  brain_join(dk) |>
  reposition_brain(hemi ~ side) |>
  ggplot(aes(fill = p)) +
  geom_sf(show.legend = FALSE) +
  facet_wrap(~groups) +
  ggtitle("correct facetting")
```
