---
title: "Plotting brain mesh atlases in plotly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting brain mesh atlases in plotly}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE)
devtools::load_all(".")
library(magrittr)
library(dplyr)
library(tidyr)
```

## Introduction
Some times, you want to plot the 3d-surfaces from freesurfer. Here, it is easier to use data from freesurfer (like annot and colorlut files) to create colours for the vectors. 
The data is somewhat more complex than the 2d ggplot polygon version `ggseg`
`ggseg3d()` will create a plotly plot, which is interactive, and provides another type of flexibility to the user. 
A lot of credit goes to A.M.Winkler and his [Brainder](https://brainder.org/) work, which supplied us with the first examples of going from `.srf` to `.ply` files, and whose scripts massively aided us in making this work.

## Basic use
The function `ggseg3d()`, is based in the plotly, it is recommended to get a little familiarized with with [plotly](https://plotly-book.cpsievert.me/). 

Out-of-the-box, ggseg() works without supplying any extra information. 
It will create a base plot of the aparc (dkt) brain segmentations.
All `[...]_3d` atlases have a built in `colour` column for default colour plotting of the segments.

```{r start, eval=F}
library(ggseg)
library(magrittr)
library(dplyr)
library(tidyr)
```
```{r starte, message=FALSE}
ggseg3d()
```


## Atlas data
The data is stored in tibbles, and looks like so:
```{r,fig.width=5}
dkt_3d
```

To grab all the data for a surface and hemisphere, you should reduce the data to one line, and then `unnest()`
```{r}
dkt_3d %>% 
  filter(surf == "inflated" & hemi == "right") %>% 
  unnest()
```

## External data supply
Particularly notice the `mesh` column, which is a list column of lists. In there is all the 6 vectors needed to create the mesh of the tri-surface plot. 
You'll also need to notive the `label`, `annot` and `area` columns, which are likely the columns you will be matching on when proviging with your own data for colours. 
You need to be meticulous when fixing your data, be sure it matches.
The function _should_ give you a warning if it's struggling to match something. 

The column you want to use for colour, needs to be supplied to the `colour` option, and you'll likely want to supply it to the `text` option, as this will add another line to the plotly hover information.

```{r,fig.width=5}
someData = dkt_3d %>% 
  filter(surf == "inflated" & hemi == "right") %>% 
  unnest() %>% 
  select(area) %>% 
  na.omit() %>% 
  mutate(p = sample(seq(0,.5, length.out = 100 ), nrow(.)) %>% 
           round(2)) 
  
  ggseg3d(.data = someData, atlas = dkt_3d, colour = "p", text = "p")
```

## Colours
<!-- You can also change the palette. Choose from ALL the paletteer palettes, just use their name. -->
<!-- ```{r,fig.width=5} -->
<!--   ggseg3d(.data = someData, atlas = dkt_3d, colour = "p", text = "p", palette = "berlin") -->
<!--   ggseg3d(.data = someData, atlas = dkt_3d, colour = "p", text = "p", palette = "RdBu") -->
<!--   ggseg3d(.data = someData, atlas = dkt_3d, colour = "p", text = "p", palette = "GrandBudapest1") -->
<!-- ``` -->

You can provide custom colour palettes either in hex or R-names
```{r,fig.width=5, eval=FALSE}
  ggseg3d(.data = someData, atlas = dkt_3d, 
          colour = "p", text = "p", 
          palette = c("forestgreen", "white", "firebrick"))
  ggseg3d(.data = someData, atlas = dkt_3d, 
          colour = "p", text = "p", palette = c("#ff0000", "#00ff00", "#0000ff"))
```

If you want to alter the colour of `NA` regions, supply `na.colour`
```{r,fig.width=5, eval=FALSE}
  ggseg3d(.data = someData, atlas = dkt_3d, 
          colour = "p", text = "p", 
          na.colour = "black", camera="medial")
  ggseg3d(.data = someData, atlas = dkt_3d, 
          colour = "p", text = "p", 
          na.colour = "#ffff00", camera="medial")
```

If you are plotting the sub-cortical structures, you might want to reduce the opacity of the `NA` structures, so that you can see the more medial structures. you may also want to add the `glassbrain` option, any value above 0 and up to 1 increases the opacity of a glass brain overlay.
```{r,fig.width=5, warning=F, message=FALSE}
somData_aseg = aseg_3d %>% 
  unnest() %>% 
  select(label) %>% 
  filter(!grepl("Ventricle|Putamen", label)) %>% 
  mutate(p = seq(1, nrow(.))) 

  ggseg3d(.data = somData_aseg, atlas = aseg_3d, 
          colour = "p", text = "p", 
          na.alpha= .5)
```

## Camera
You can rotate away all you like. if you want to start from the medial view, you can use the `camera` option
```{r,fig.width=5, eval=FALSE}
  ggseg3d(.data = someData, atlas=dkt_3d, colour = "p", camera = "medial")
```

There is also a surface of the white-matter
```{r,fig.width=5, eval=FALSE}
 ggseg3d(atlas=dkt_3d, surface="white")
```

There is much more you can do with the camera options, or by adding extra layout options to plotly. 
You may want to have a look at these resources if you want to do more complex adaptations to the plots:

https://plot.ly/r/axes/#modifying-axes-for-3d-plots  
https://plot.ly/r/trisurf/  
https://moderndata.plot.ly/trisurf-plots-in-r-using-plotly/




